---
title: "Confidence Ellipse"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Confidence Ellipse}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Confidence ellipses are extension of the concept of a confidence interval, which is used for a single variable. The ellipse is centered at the point representing the sample mean values of the two variables. Its size and shape are determined by the chosen confidence level (e.g., 95%) and the covariance matrix.

```{r message=FALSE, warning=FALSE}
library(magrittr)
library(dplyr)
library(ggplot2)
library(patchwork)
```

```{r message=FALSE, warning=FALSE}
library(ConfidenceEllipse)
```

```{r}
data(glass, package = "ConfidenceEllipse")
```

# Coordinate points

The `confidence_ellipse` function is used to compute the coordinate points of the confidence ellipse and then the ellipse is plotted on a two-dimensional plot `x` and `y` of the data. The ellipse represents a region that, with a specified confidence level `conf_level`, is expected to contain the true mean of the bivariate distribution.

```{r message=FALSE, warning=FALSE}
ellipse99 <- confidence_ellipse(glass, x = SiO2, y = Al2O3, conf_level = 0.99)
ellipse95 <- confidence_ellipse(glass, x = SiO2, y = Al2O3, conf_level = 0.95)
ellipse90 <- confidence_ellipse(glass, x = SiO2, y = Al2O3, conf_level = 0.90)
```

```{r}
ellipse99_rob <- confidence_ellipse(glass, x = SiO2, y = Al2O3, conf_level = 0.99, robust = TRUE)
ellipse95_rob <- confidence_ellipse(glass, x = SiO2, y = Al2O3, conf_level = 0.95, robust = TRUE)
ellipse90_rob <- confidence_ellipse(glass, x = SiO2, y = Al2O3, conf_level = 0.90, robust = TRUE)
```

```{r message=FALSE, warning=FALSE}
ellipse99 %>% glimpse()
```

```{r}
ggplot() +
  geom_path(data = ellipse99, aes(x = x, y = y), color = "red", linewidth = 0.7, linetype = "dashed") +
  geom_path(data = ellipse95, aes(x = x, y = y), color = "blue", linewidth = 0.7, linetype = "dashed") +
  geom_path(data = ellipse90, aes(x = x, y = y), color = "green", linewidth = 0.7, linetype = "dashed") +
  geom_path(data = ellipse99_rob, aes(x = x, y = y), color = "red", linewidth = 1L, linetype = "solid") +
  geom_path(data = ellipse95_rob, aes(x = x, y = y), color = "blue", linewidth = 1L, linetype = "solid") +
  geom_path(data = ellipse90_rob, aes(x = x, y = y), color = "green", linewidth = 1L, linetype = "solid") +
  geom_point(data = glass, aes(x = SiO2, y = Al2O3), fill = "grey", color = "black", shape = 21, size = 3L) +
  annotate("text", x = 70, y = 3.5, label = "Classical confidence ellipses", color = "black", size = 4) +
  annotate("text", x = 74, y = 2.3, label = "Robust confidence ellipses", color = "black", size = 4) +
  scale_color_brewer(palette = "Set1", direction = 1) +
  xlim(50, 80) +
  ylim(-.5, 5) +
  labs(
    x = "SiO2 (wt.%)",
    y = "Al2O3 (wt.%)",
    title = "Confidence ellipses",
    caption = "
    Figure 1: Comparison between classical and robust confidence ellipse."
  ) +
  theme_bw() +
  theme(
    aspect.ratio = .7,
    panel.grid = element_blank(),
    legend.position = "none"
  )
```

# Group-based confidence ellipses

For grouping bivariate data, the `.group_by` argument can be used if the data contains an unique grouping variable (`.group_by = NULL` by default). When a grouping variable is provided, the function will compute the ellipses separately for each level of the factor. It's important to note that the grouping variable should be appropriately coded as a factor before passing it to the `.group_by` argument. If the variable is currently stored as a character or numeric type, you may need to convert it to a factor using functions like `as.factor()` or `forcats::as_factor()`.

The function includes a `distribution` argument that allows users to choose between two statistical approaches for ellipse calculation:

-   "normal" (default): Uses the chi-square distribution quantile
-   "hotelling": Uses Hotelling's T² distribution quantile

```{r message=FALSE, warning=FALSE}
ellipse_grp <- confidence_ellipse(glass, x = SiO2, y = Na2O, .group_by = glassType)
ellipse_grph <- confidence_ellipse(glass, x = SiO2, y = Na2O, .group_by = glassType, distribution = "hotelling")
```

```{r}
ellipse_grp %>% print()
```

```{r}
ellipse_grph %>% print()
```

```{r}
p1 <- ggplot() +
  geom_polygon(data = ellipse_grp, aes(x = x, y = y, fill = glassType), color = "black", alpha = .45) +
  geom_point(data = glass, aes(x = SiO2, y = Na2O, color = glassType, shape = glassType), size = 3L) +
  scale_fill_brewer(palette = "Set1", direction = 1) +
  scale_color_brewer(palette = "Set1", direction = 1) +
  labs(
    x = "SiO2 (wt.%)",
    y = "Na2O (wt.%)"
  ) +
  theme_bw() +
  theme(
    panel.grid = element_blank(),
    legend.position = "none"
  )
```

```{r}
p2 <- ggplot() +
  geom_polygon(data = ellipse_grph, aes(x = x, y = y, fill = glassType), color = "black", alpha = .45) +
  geom_point(data = glass, aes(x = SiO2, y = Na2O, color = glassType, shape = glassType), size = 3L) +
  scale_fill_brewer(palette = "Set1", direction = 1) +
  scale_color_brewer(palette = "Set1", direction = 1) +
  labs(
    x = "SiO2 (wt.%)",
    y = "Na2O (wt.%)",
    caption = "
    Figure 2: Comparison between normal distribution (left) and Hotelling's T² distribution (right) ellipses."
  ) +
  theme_bw() +
  theme(
    panel.grid = element_blank(),
    legend.position = "none"
  )
```

```{r}
patchwork::wrap_plots(p1, p2, ncol = 2) +
  patchwork::plot_annotation(
    title = "Statistically-based confidence ellipses drawn around each group",
    theme = theme(plot.title = element_text(hjust = 0.5))
  )
```

As shown in the comparative plots, the difference between these approaches can be substantial, especially for smaller sample sizes. The Hotelling's T² ellipses provide more realistic confidence regions that properly account for the fact that we're estimating population parameters from sample data. As the sample size increases the differences between both approaches diminish.


