---
title: "Confidence Ellipse"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Confidence Ellipse}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

A confidence ellipse is a graphical representation that shows the variability or uncertainty in estimates of the mean values of two normally distributed variables in a bivariate dataset. It is an extension of the concept of a confidence interval, which is used for a single variable. The ellipse is centered at the point representing the sample mean values of the two variables. Its size and shape are determined by the chosen confidence level (e.g., 95%) and the covariance matrix of the variables.

```{r message=FALSE, warning=FALSE}
library(magrittr)
library(dplyr)
library(ggplot2)
```

```{r message=FALSE, warning=FALSE}
library(ConfidenceEllipse)
```

```{r}
data(glass, package = "ConfidenceEllipse")
```

# Coordinate points

The `confidence_ellipse` function is used to compute the coordinate points of the confidence ellipse and then the ellipse is plotted on a two-dimensional plot `x` and `y` of the data. Points that lie within the ellipse are considered to be part of the underlying distribution with the specified confidence level `conf_level`.

```{r message=FALSE, warning=FALSE}
ellipse_99 <- confidence_ellipse(glass, x = MgO, y = Cl2O, conf_level = 0.99)
ellipse_95 <- confidence_ellipse(glass, x = MgO, y = Cl2O, conf_level = 0.95)
ellipse_90 <- confidence_ellipse(glass, x = MgO, y = Cl2O, conf_level = 0.90)
```

```{r message=FALSE, warning=FALSE}
ellipse_99 %>% glimpse()
```

```{r}
ggplot() +
  geom_point(data = ellipse_99, aes(x = x, y = y), color = "red", size = .1) +
  geom_point(data = ellipse_95, aes(x = x, y = y), color = "blue", size = .1) +
  geom_point(data = ellipse_90, aes(x = x, y = y), color = "green", size = .1) +
  geom_point(data = glass, aes(x = MgO, y = Cl2O), color = "black", size = 3L) +
  xlim(-.5, 6) +
  ylim(-0.05, 1.5) +
  scale_color_brewer(palette = "Set1", direction = 1) +
  labs(x = "MgO (wt.%)", y = "Cl2O (wt.%)") +
  theme_bw() +
  theme(legend.position = "none")
```

# Grouping

For grouping bivariate data, the `.group_by` argument can be used if the data contains an unique grouping variable (`.group_by = NULL` by default). When a grouping variable is provided, the function will compute the ellipses separately for each level of the factor. It's important to note that the grouping variable should be appropriately coded as a factor before passing it to the `.group_by` argument. If the variable is currently stored as a character or numeric type, you may need to convert it to a factor using functions like `as.factor()` or `forcats::as_factor()`.

```{r message=FALSE, warning=FALSE}
ellipse_grp <- confidence_ellipse(glass, x = MgO, y = Cl2O, .group_by = glassType)
```

```{r}
ellipse_grp %>% glimpse()
```

```{r}
ggplot() +
  geom_point(data = glass, aes(x = MgO, y = Cl2O, colour = glassType, shape = glassType), size = 3L) +
  geom_line(data = ellipse_grp, aes(x = x, y = y, colour = glassType), linewidth = .1) +
  xlim(0, 6) +
  ylim(-0.05, 1.5) +
  scale_color_brewer(palette = "Set1", direction = 1) +
  labs(x = "MgO (wt.%)", y = "Cl2O (wt.%)") +
  theme_bw() +
  theme(legend.position = "none")
```
